Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-05-30 jhir/latex-manuscript/manuscript.tex
update manuscript with reviewer's feedback

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{geometry}
\usepackage{pdflscape}
%
\usepackage{mathptmx}      % use Times fonts if available on your TeX system

@@

HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, a HMM-based classifier was \hl{applied to Doppler} ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}. In another study, HMM was used to capture important
characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.

@@

\subsection{Dataset}
\label{subsec:dataset}
The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}. The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied frequent pattern mining to identify frequently occurring patterns of temporally ordered patient-counselor behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular non-close frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}. A \textit{frequent pattern} is defined as a pattern of temporally ordered observations, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}. For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern is closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed \hl{prior work}~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences. For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{geometry}
\usepackage{pdflscape}
%
\usepackage{mathptmx}      % use Times fonts if available on your TeX system

@@

HMMs are widely used for the analysis of sequence data due to their ability to model \hl{long-range dependencies} between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden state'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, a HMM-based classifier was \hl{applied to Doppler} ultrasound imaging data to extract features from from the images that were then used to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}. In another study, HMM was used to capture important
characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states and the sets of related behavior codes associated with the hidden states, may correspond a patient’s underlying motivational state during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (a special class of change talk where patients express their intentions, plans and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patients' behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that are likely to lead to patient change talk and commitment language.

@@

\subsection{Dataset}
\label{subsec:dataset}
The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding greetings, farewell and conversations for interview setup such as camera and table setup.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied frequent pattern mining to identify frequently occurring \hl{patterns of patient-counselor} behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art close frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular non-close frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}. A \textit{frequent pattern} is defined as a \hl{pattern of observations}, which appears in a given set of sequences more often than a user-specified threshold called the minimum \textit{support count}. For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern is closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns, since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns, since none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed \hl{prior work}~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences. For each pattern, the statistical significance of difference between pattern occurrence in successful and unsuccessful communications was reported by computing p-value using Pearson's chi-square test.
